---
title: "Auckland Bike Counter Analysis"
author: "kimnewzealand"
date: "15 October 2018"
output: html_document
---

## Introduction

The motivations for cycling around town can be for fitness, economic or environmental reasons. However in Auckland the weather and [the 53 odd volcanoes](https://en.wikipedia.org/wiki/Auckland_volcanic_field) could be deterrents to would-be cyclists.

Auckland Transport (AT) publishes daily and monthly bike count data from counters around Auckland.

We will analyse and visualise this data, then fit a model to forecast Auckland bike activity and trends to see if cyclists are braving the weather and taking on the hills.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load Packages

```{r packages, message=FALSE, warning=FALSE}
# Load the tidyverse metapackage
library(tidyverse)
# Load skimr for summary statistics
library(skimr)
# Load the leaflet and htmltools packages for maps
library(leaflet)
library(htmltools)
# Load the sf package for map coordinate manipulations
library(sf)
# Load the lubridate package for time variable manipulations
library(lubridate)
# Load the forecast package for time series
library(forecast)
# Load the cowplot package for side by side ggplots
library(cowplot)
```


## Data

**Import data**

Use the read_csv functions from the [readr](https://cran.r-project.org/web/packages/readr/README.html) R package to import the csv files directly from urls. These function can be used for most csv files, reading and parsing the files to dataframes.

```{r import data, warning=FALSE}
# Import the bikes daily counter dataset for August
url <- "https://at.govt.nz/media/1977986/august2018akldcyclecounterdata.csv"
bikes_daily <- read_csv(url)
# Import the bikes mounthly counter dataset. Set the counter variables types to numeric
url2 <-"https://at.govt.nz/media/1975708/monthlyakldcyclecountdatanov2010-dec2017.csv"
bikes_monthly <- read_csv(url2, 
                          col_types = cols(.default = col_number(), Month=col_character()))
```

**Data Summaries**

View the summary statistics with the [skimr](https://cran.r-project.org/web/packages/skimr/) R package. 

```{r skim bikes_daily}
# Produce a skim summary excluding the histograms as these do not plot in R markdown currently
skim_with(numeric = list(hist = NULL))
skim_with(
  numeric = list(p0 = NULL, p25 = NULL, p50 = NULL, p75 = NULL, p100 = NULL,sd = NULL,hist = NULL),
  integer = list(p0 = NULL, p25 = NULL, p50 = NULL, p75 = NULL, p100 = NULL,sd = NULL,hist = NULL)  
)
skim(bikes_daily)
```

The `bikes_daily` dataset has the following variables in wide format:  

**Date** : daily from `r bikes_daily[1,1]`  
**Bike counter street location names** : `r dim(bikes_daily)[2]-1` locations  

It seems like the last row consists of NAs, so we can remove this row in the cleaning step.

```{r skim bikes_monthly}
# Produce a skim summary excluding the histograms as these do not plot in R markdown currently
skim(bikes_monthly)
```

The `bikes_monthly` dataset has the following variables in wide format:  

**Date** : monthly from `r bikes_monthly$Month[2]`  
**Bike counter street location Names** : `r dim(bikes_monthly)[2]-1` locations

There are missing values as NA's to be considered in data aggregations.

**Auckland Population Data**

Population estimates for regional council, territorial authority (including Auckland local board), area unit, and urban areas is available from [DataHub table viewer](http://nzdotstat.stats.govt.nz/wbos/index.aspx). In order to download the data, a request needs to be submitted online but the Stats NZ link doesn't currently work. As an interim workaround, a manual dataframe is created based on the online query results.

```{r create auckland pop dataset}
# Create Auckland population dataset
population <- data.frame(Date= c("30-June-11","30-June-12","30-June-13","30-June-14","30-June-15",         "30-June-16","30-June-17"), Auckland = c(1459600, 1476500, 1493200, 1526900, 1569900, 1614500, 1657200   ))
```


## Data Cleaning

The `bikes_daily` and `bikes_monthly` counter data do not have coordinates available so we will manually create a dataframe with selected points with the longitude and latitude coordinates from a [GPS coordinates lookup website](https://gps-coordinates.org/new-zealand-latitude.php). 

- Curran_Street  
- Quay_Street  
- Grafton_Gully  

Feedback has been sent to AT to add the longitude and latitude coordinates to their csv files, so that further mapping and joins with other data can be performed another time.

```{r create streets coords}
# Create a streets datafarme
streets <- data.frame(street =c("Curran_Street","Quay_Street","Grafton_Gully"), longitude=c(174.739289,174.773376,174.7678),latitude=c(-36.840521,-36.844713,-36.8623), stringsAsFactors = FALSE)
```


Now clean the the `bikes_daily` and `bikes_monthly` using [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) and [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html) R packages.

```{r cleaning count datasets}
# Convert the character date variable to a date variable using lubridate
bikes_daily$Date <- dmy(bikes_daily$Date)
# Add a new weekday variable using lubridate and dplyr
bikes_daily <- bikes_daily %>% 
      mutate(weekday =weekdays(Date)) %>% 
      # Remove the last row with NAs in the dataframe using slice using this tip https://stackoverflow.com/questions/38916660/r-delete-last-row-in-dataframe-for-each-group
      slice(-n())

# View the column names by position in bikes_daily to check they are the 3 points we have selected
names(bikes_daily)[4]
names(bikes_daily)[9]
names(bikes_daily)[29]

# Rename the 3 points in bikes_daily to match the streets points
names(bikes_daily)[4] <- streets[1,1]
names(bikes_daily)[9] <- streets[2,1]
names(bikes_daily)[29] <- streets[3,1]

# View the column names by position in bikes_monthly to check they are the 3 points we have selected
names(bikes_monthly)[5]
names(bikes_monthly)[11]
names(bikes_monthly)[32]

# Rename the 3 points in bikes_monthly to match the streets points
names(bikes_monthly)[5] <- streets[1,1]
names(bikes_monthly)[11] <- streets[2,1]
names(bikes_monthly)[32] <- streets[3,1]

# Add the row sums ignoring the NAs and remove the first row which has a month as Te Atatu
bikes_monthly <- bikes_monthly %>% 
      mutate(sumStreets = rowSums(.[2:dim(bikes_monthly)[2]],na.rm = TRUE)) %>% 
      slice(-1)
# Convert the character date variable to a date variable using lubridate
bikes_monthly$Month <- dmy(paste0("01-", bikes_monthly$Month)) 
# In order to select the bikes columns as the vector of points in the streets dataframe, we need to provide column names as variables which contain strings. We could use select_ as the standard evaluation counterpart of select. However from the docs select_ has been deprecated in favour of tidy eval. https://timmastny.rbind.io/blog/nse-tidy-eval-dplyr-leadr/#expressions. We use the unquote !! https://adv-r.hadley.nz/quasiquotation.html to tell dplyr to select the variables that match the values in the streets$street vector and combine with the Date and weekday variable in the bikes dataset
bikes_subset <- bikes_daily %>% 
      select(!!streets$street) %>% 
      cbind(bikes_daily %>% select(Date,weekday))

# Now convert this dataframe into a tidy dataset using gather, where the Key is the street and the value is the count http://garrettgman.github.io/tidying/
bikes_subset <- bikes_subset %>% 
      gather(key = "street", value="counts",1:3)

# Now join the streets coordinates and the bikes subset dataframe using dplyr function inner_join    
bikes_streets <- streets %>%
      inner_join(bikes_subset,by = "street")
```

```{r create sf object}
# Creat a sf object from the streets dataframe with points geometries with crs as 2193 https://epsg.io/2193
bikes_streets_sf <- st_as_sf(bikes_streets, 
                             coords = c("longitude", "latitude"), crs=2193)
```

# Visualisations

**Bike Counter Locations Daily Data**

Let's visualise the selected bike counter locations using the [leaflet](https://cran.r-project.org/web/packages/leaflet/) R package.

```{r leaflet, warning=FALSE}
# Create a leaflet map centred at the mean of the streets points
long <- mean(streets$longitude)
lat <- mean(streets$latitude)
m <- leaflet() %>% 
      setView(lng = long, lat = lat, zoom = 13) %>% 
      addTiles()
m %>% 
      addMarkers(data = bikes_streets_sf, 
                 popup = ~htmlEscape(street)) %>% 
      addProviderTiles(providers$ Stamen.TonerBackground)
```

Let's view the daily counts by the 3 counter points with scatter plot and labelled scatterplot, with the regression trend line.

```{r weekday plots, message=FALSE, warning=FALSE}
# Plot a facetted plot of the  daily bike counts
ggplot(bikes_streets,aes(Date,counts))+ 
      geom_point() +
      geom_smooth(method="lm") +
      facet_wrap(~street) +
      theme_bw()+
      ggtitle("Daily Bicycle Counts and Weekdays for Counter Locations")
```

```{r weekday label plots, message=FALSE, warning=FALSE}
# Plot a facetted plot of the  daily bike counts
ggplot(bikes_streets,aes(Date,counts))+ 
      geom_point() +
      geom_label(aes(label=weekday)) +
      geom_smooth(method="lm") +
      facet_wrap(~street) +
      theme_bw()+
      ggtitle("Daily Bicycle Counts and Weekdays for Counter Locations")
```

  
It seems like there is a downward trend for August, perhaps this is related to the winter weather in August?
  
Next create a ggplot of the total aggregated count by weekday and street.

```{r weekdays bar plots}
# Create a gpplot of the total aggregated count by weekday and street
bikes_streets %>% 
      group_by(weekday,street) %>% 
      summarise(weekday_sum=sum(counts)) %>% 
      na.omit() %>% 
      # Plot the weekday sum of counts and order the factor levels of the weekday variable but then use forcats fc_rev to reverse the ordering of labels when we add the coord_flip
      ggplot(aes(forcats::fct_rev(factor(weekday, c("Monday", "Tuesday", "Wednesday", "Thursday","Friday", "Saturday","Sunday"))) , weekday_sum, fill=street))  + 
      geom_col(position="dodge") +
      scale_fill_viridis_d()  +     
      xlab("Weekday")+
      ylab("Sum of Weekday Counts") +
      coord_flip()  +
      theme_bw() +
      ggtitle("Total Aggregated Bicyle Count by Weekday and Street")+ 
      labs(caption = "Source: AT https://at.govt.nz/cycling-walking/research-monitoring/")
```

Weekdays notably Thursday appears to be the most popular day in Grafton Gully and Quay Street, but Sunday appears the most popular day for cycling in Curran Street.


**Bike Counter Locations Monthly Aggregated Data**

Now lets turn to the monthly bike count dataset `bike_monthly` and look at this as a time series. 

First we will use the base decompose function to break the time series into its components.

```{r decompose monthly counts}
# Create a time series object bikes_monthly_ts
bikes_monthly_ts <- ts(bikes_monthly$sumStreets, start=c(2010,11), frequency=12)
decomp <- decompose(bikes_monthly_ts)
# Plot the decomposed time series
autoplot(decomp)
```

From the decomposed components it appears there is a repetitive seasonality pattern dipping in winter months and an overall upwards trend.

```{r pop and bike plots}
# Plot the Auckland population
population$Date <- dmy(population$Date)
pop_ts <- ts(population$Auckland, start=c(2011), frequency=1)

bike_trend <- decomp$trend
# Plot the Auckland bicycle growth
bike_plot <- autoplot(bike_trend,na.rm=TRUE) +
      ylab("Bicycle Counts") +
      xlab("Date")+
      scale_y_continuous(labels = scales::comma)+
      ggtitle("Auckland Estimated Bicyle Counts") +
      theme_bw() + 
      labs(caption = "Source: AT https://at.govt.nz/cycling-walking/research-monitoring/")
# Plot the Auckland population growth
pop_plot <- autoplot(pop_ts)+xlab("Date") +
      ylab("Population number")+
      scale_y_continuous(labels = scales::comma)+
      ggtitle("Auckland Estimated Population") +
      theme_bw()+ 
      labs(caption = "Source: NZ Stats http://nzdotstat.stats.govt.nz/")
# Plot side by side plots with cowplot package
plot_grid(bike_plot,pop_plot, labels = "AUTO")
# save_plot("cowplot.png", p, ncol = 2)
```


The cycling trend in A looks remarkably similar to Auckland population growth in B?!? Are more people cycling regardless of the hilly roads ?


Finally create an arima model using the auto.arima function from the [forecast](https://cran.r-project.org/web/packages/forecast/index.html) R package.

```{r fit arima model}
# Fit an arima model using auto.arima function
fit <- bikes_monthly_ts %>% 
      auto.arima(seasonal=TRUE) 
# Produce a summary of the fitted model
summary(fit)
# Check ACF and PACF plots for model residuals
tsdisplay(residuals(fit), lag.max=45, main='(0,1,1) Model Residuals')
# Plot the forecasted model
fit %>% 
      forecast() %>% 
      autoplot() +
      ylab("Bicycle counts")+
      xlab("Date")+
      scale_y_continuous(labels = scales::comma)+
      scale_x_continuous(labels = scales::number)+
      theme_bw()
```

This model is a [seasonal](http://people.duke.edu/~rnau/seasarim.htm) Arima model. The moving average model has  p=0 autoregressive terms, d=1 , q=1 moving average terms, where d=1 ie differencing of order 1. The seasonal part of the model has P=2 seasonal autoregressive (SAR) terms, D=1 number of seasonal differences, Q=0 seasonal moving average (SMA) terms.

## Conclusion

The bike count data is interesting data to look at, especially the time series components forecasting model. Once the coordinates are available further comparisons could be made because the locations and over time.
